Tools for identifying unexpectedly low microfossil count sums

### 2020-06-15 Rmd/count_check_MS.Rmd
wildhorse

loadd(bird_singleton_summary)
loadd(testate_summ)
loadd(last_chance_summ)
loadd(pollen_summ, hershop_pollen_summ, pollenA, pollen1234)
loadd(owens_summ1, owens_summ, owens_est_summ)
loadd(diatom1_summ, diatom2_summ, chironomid1_summ, marine1_summ, chironomid2_summ)

@@

The pollen data from the Neotoma database included over `r pollen_summ$n_assemblages` assemblages in `r pollen_summ$n_datasets %>% format()` datasets. 
Nearly all assemblages have singletons (`r pollen_summ$p_singletons`%) and a greatest common divisor of one (`r pollen_summ$p_gcd1`%). 
The assemblages with a greatest common divisor above one are not randomly distributed among the datasets: only `r pollen_summ$dataset_gcd1`% of the datasets have any such assemblages.
Some datasets with a high proportion of assemblages with a greatest common divisor above one are older datasets which may, despite the available metadata, have been back-transformed from percent data or digitised with the loss of rare taxa and precision.
For example, all but `r hershop_pollen_summ$has_singleton %>% word()` of the `r hershop_pollen_summ$n` assemblages in the archived pollen data from Hershop Bog, Texas have a greatest common divisor of two or more (the exception has a single singleton and is otherwise divisible by two). 
The original data [@Patty1968] were counted to exactly 200 pollen grains and fern spores. 
It seems likely that the data were converted to percent, rounded and then back-tranformed to counts.
For most purposes, such back-transformed or digitised data are perfectly adequate.

Excluding such datasets would further increase the proportion of assemblages with singletons. 
Some other Neotoma datasets are discussed in section \@ref(unexpected-count-data) of this paper.

@@

Some of the pollen datasets in the Neotoma database have an inexplicably high proportion of assemblages with a greatest common divisor greater than one. 
Here I focus on five anonymised datasets.

@@

Dataset pollen3, published includes `r pollen3$n_samp` assemblages with a median richness of `r pollen3$taxa_median` taxa.
Count sums vary between `r pollen3$count_min` and `r pollen3$count_max`.
The greatest common divisor is three for all assemblages: all `r pollen3$n_count` counts in the dataset are divisible by three. 
Assuming (optimistically) the distribution of counts is (at least locally) uniform, the probability of _n_ counts being divisible by _k_ is $k^{-n}$.
For a dataset this large, the probability of having all counts divisible by three by chance is 10^`r ceiling(- pollen3$n_count * log10(3))`^.
Given this very small probability, it seems reasonable to conclude that the archived data are not the original counts.

>>>>

loadd(bird_singleton_summary)
loadd(testate_summ)
loadd(last_chance_summ)
loadd(pollen_summ, hershop_pollen_summ, wildhorse, pollen_summ1, pollen_summ2, pollen3, pollen1234)
loadd(owens_summ1, owens_summ, owens_est_summ)
loadd(diatom1_summ, diatom2_summ, chironomid1_summ, marine1_summ, chironomid2_summ)

@@

The pollen data from the Neotoma database included over `r pollen_summ$n_assemblages` assemblages in `r pollen_summ$n_datasets %>% format()` datasets. 
Conforming to expectations, nearly all assemblages have singletons (`r pollen_summ$p_singletons`%) and a greatest common divisor of one (`r pollen_summ$p_gcd1`%). 
The assemblages without singletons or with a greatest common divisor above one are not randomly distributed among the datasets.
Half of the assemblages without singletons are in `r pollen_summ$half_no_sing` datasets.
`r Words(pollen_summ$half_gcd)` datasets contain half of the `r pollen_summ$n_gcd1` assemblages with a greatest common divisor above one and only `r pollen_summ$dataset_gcd1`% of the datasets have any such assemblages.
Some Neotoma datasets are with a high proportion of assemblages without singletons or a with a greatest common divisor above one are discussed in section \@ref(unexpected-count-data) of this paper.

@@

This section explores some of the pollen datasets in the Neotoma database that have a high proportion of assemblages without singletons or with a greatest common divisor greater than one.

Some of these datasets are older datasets which may, despite the available metadata, have been back-transformed from percent data or digitised with the loss of rare taxa and precision.
For example, all but `r hershop_pollen_summ$has_singleton %>% word()` of the `r hershop_pollen_summ$n` assemblages in the archived pollen data from Hershop Bog, Texas have a greatest common divisor of two or more (the exception has a single singleton and is otherwise divisible by two). 
The original data [@Patty1968] were counted to exactly 200 pollen grains and fern spores. 
It seems likely that the data were converted to percent, rounded and then back-transformed to counts.
For most purposes, such back-transformed or digitised data are perfectly adequate.

Some datasets with assemblages without singletons have few taxa.
For example, the `r wildhorse$n_assemblage` assemblage dataset from Wildhorse Lake [@ref] includes only `r wildhorse$n_taxa_dataset` taxa (`r mean(pollen_summ2$n_taxa_dataset > wildhorse$n_taxa_dataset) * 100` % of neotoma pollen datasets have more than this), and relatively large count sums (median `r wildhorse$median_count`). 
While `r wildhorse$no_singletons * 100`% of assemblages in this dataset lack singletons, even more assemblage (`r wildhorse$no_zeros * 100`%) lack taxa with zero counts. 
This is in contrast to most (`r pollen_summ$has_zeros * 100`%) neotoma pollen assemblages which have zero counts.  
These patterns could arise either when there is a high sampling effort in a low diversity site, or when the archived data do not include the entire pollen assemblage, instead focusing on the subset of abundant taxa of most interest to the authors.

For some other datasets, the reason that assemblages lack singletons or have a greatest common divisor above one is less clear. 
These are treated anonymously.

@@

In dataset pollen3, the greatest common divisor is three for all assemblages: all `r pollen3$n_count` counts in the dataset are divisible by three (Table \@ref(tab:unexpected-table))). 
Assuming (optimistically) the distribution of counts is (at least locally) uniform, the probability of _n_ counts being divisible by _k_ is $k^{-n}$.
For a dataset this large, the probability of having all counts divisible by three by chance is 10^`r ceiling(- pollen3$n_count * log10(3))`^.
Given this very small probability, it seems reasonable to conclude that the archived data are not the original counts.
